Stochastic Approximation Monte Carlo with Dynamic Update Factor (SAD)

### 2018-08-22 papers/histogram/sad.tex
cut one line from equations
made changes to convergence section for SAD.

We can place a rigorous \emph{lower} bound $t_{\min}$ on the number of
moves required to find the true energy by considering the total change
that needs to be made to the entropy.
\begin{align}
  \Delta S_{\text{tot}} \equiv \sum_E S(E) - S_{\min}
\end{align}
The minimum number of moves that could converge the entropy is
the number of moves that will enable the above total entropy change,
which we can approximate using an integral:
\begin{align}
   \Delta S_{\text{tot}} &= \sum_{t=0}^{t_{\min}} \gamma_t \\
  &= \int_0^{t_{\min}} \gamma(t)dt \\
  &= t_0 + \int_{t_0}^{t_{\min}} \frac{t_0}{t}dt
  \\
  &= t_0\left(1 + \ln\left(\frac{t_{\min}}{t_0}\right)\right)
\end{align}
Solving for ${t_{\min}}$ we find that
\begin{align}
  {t_{\min}} &= t_0 e^{\frac{\Delta S_{\text{tot}}}{t_0} - 1}
\end{align}
which means that the minimum time to convergence grows exponentially
as $t_0$ is made smaller.  You \emph{seriously} don't want to underestimate
$t_0$!

@@

\subsection{Convergence}
{\color{red}
It is for this reason that we propose a way to determine the update factor
``dynamically'' while still satisfying the criteria for convergence.
For the update factor to scale well across both discrete and continuous systems,
it must include the number of states $N_S$. We define the average entropy as
$\bar S$ and the time since the simulation last found a new energy state $t_L$.
We then can write the following expression for the \emph{proposed} update factor.
\begin{align}
\gamma = N_S \frac{\bar S + \nicefrac{t}{t_L}}{N_{S}^2 + \nicefrac{t^2}{t_L}}
\end{align}
We now consider the following limiting cases:
\subsubsection{$t \propto N_S \approx t_L$ and $ t << N_S$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{\bar S}{N_S} \approx \frac{\bar S}{t}
\end{align}
Because $t \propto N_S$, the update factor scales as 1/t which is consistent
with the requirement for SAMC convergence.
\subsubsection{$t >> N_{S}^2$ and $ t\approx t_L$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{N_S \bar S t_L}{t^2} \approx \frac{N_S \bar S}{t}
\end{align}
\subsubsection{$t >> t_L \bar S$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{N_S}{t}
\end{align}
We now integrate (sum) the update factor from the last time an energy state was found
to the final simulation time to ensure that convergence is guaranteed.
\begin{align}
\int_{t_L}^{t_f} \gamma dt \approx t_L N_S \left[ \int_{t_L}^{t_f}
\frac{\bar S -1}{t^2} dt + \frac{1}{t_L}\int_{t_L}^{t_f} \frac{1}{t} dt \right]
\end{align}
From this we obtain
\begin{align}
\int_{t_L}^{t_f} \gamma dt \approx t_L N_S \left[\left(\bar S -1\right)\left(\frac{1}{t_L}-\frac{1}{t_f}\right) + \frac{1}{t_L} \ln\left(\frac{t_f}{t_L}\right) \right]
\end{align}
Simplifying, we see that as $t_f \rightarrow \infty$ then $\gamma \rightarrow \infty$ satisfying
the requirement for SAMC convergence.
\begin{align}
\gamma \approx \left(1-\frac{t_L}{t_f}\right)N_S\left(\bar S -1\right) + N_S \ln\left(\frac{t_f}{t_L}\right)
\end{align}
}

>>>>

We can place a rigorous \emph{lower} bound $t_{\min}$ on the number of
moves required to find the true energy by considering the total change
that needs to be made to the entropy.
\begin{align}
  \Delta S_{\text{tot}} \equiv \sum_E S(E) - S_{\min}
\end{align}
The minimum number of moves that could converge the entropy is
the number of moves that will enable the above total entropy change,
which we can approximate using an integral:
\begin{align}
   \Delta S_{\text{tot}} &= \sum_{t=0}^{t_{\min}} \gamma_t \\
  % &= \int_0^{t_{\min}} \gamma(t)dt \\
  &= t_0 + \int_{t_0}^{t_{\min}} \frac{t_0}{t}dt
  \\
  &= t_0\left(1 + \ln\left(\frac{t_{\min}}{t_0}\right)\right)
\end{align}
Solving for ${t_{\min}}$ we find that
\begin{align}
  {t_{\min}} &= t_0 e^{\frac{\Delta S_{\text{tot}}}{t_0} - 1}
\end{align}
which means that the minimum time to convergence grows exponentially
as $t_0$ is made smaller.  You \emph{seriously} don't want to underestimate
$t_0$!

@@

\subsection{Convergence}
{\color{red}
Although SAD determines the update factor ``dynamically'', the method must still 
satisfy the criteria for convergence. Similar to the discussion that was presented
on the convergence for SAMC, we approximate the total entropy change using an integral.
The update factor is integrated from the last time an energy has been discovered $t_L$ to
some minimum time $t_{\min}$ necessary for convergence.
\begin{align}
\Delta S_{\text{tot}} = \int_{t_L}^{t_{\min}} \gamma dt \approx t_L + t_L \ln\left(\frac{t_{\min}}{t_L}\right) \\
+ N_S t_L \Delta S_{\text{tot}} \left(\frac{1}{t_{\min}}-\frac{1}{t_L}\right) 
\end{align}

Simplifying, we see that as $t_{\min} \rightarrow \infty$ then $\gamma \rightarrow \infty$ satisfying
the requirement for SAMC convergence. We can solve for the minimum time
\begin{align}
\Delta S_{\text{tot}} = \Delta S_{\text{SAMC}} + N_S t_L \Delta S_{\text{tot}} \left(\frac{1}{t_{\min}}-\frac{1}{t_L}\right)
\end{align}
Solving for the minimum time necessary $t_{\min}$, we find that
\begin{align}
t_{\min} = t_L \left( \frac{N_S \Delta S_{\text{tot}}}{\Delta S_{\text{tot}} + \Delta S_{\text{SAMC}}} + 1\right)
\end{align}
Immediately we see that the minimum time necessary is roughly proportional
to the last time the simulation has found an energy times a ratio of entropies.

\jpsays{Are these right??? Check!}
We now consider the following limiting cases:
\subsubsection{$t \propto N_S \approx t_L$ and $ t << N_S$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{\bar S}{N_S} \approx \frac{\bar S}{t}
\end{align}
Because $t \propto N_S$, the update factor scales as 1/t which is consistent
with the requirement for SAMC convergence.
\subsubsection{$t >> N_{S}^2$ and $ t\approx t_L$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{N_S \bar S t_L}{t^2} \approx \frac{N_S \bar S}{t}
\end{align}
\subsubsection{$t >> t_L \bar S$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{N_S}{t}
\end{align}

}
